---
echo: false
---

# Derivatives {#sec-derivatives}

```{r setup_derivs} 
#| message: false
library("tidyverse")
library("cowplot")
library("dataverse")
library("ggrepel")
theme_set(theme_cowplot())
```

```{r get_pres_data} 
#| message: false
#| cache: true
df_pres_county <- get_dataframe_by_name(
  filename = "countypres_2000-2024.tab",
  dataset = "10.7910/DVN/VOQCHQ",
  server = "dataverse.harvard.edu",
  version = "15.0"
)
```

```{python}
# Set up concept dictionary
concepts = {}
```

## Motivation: How do we calculate the regression slope?

Let's play a political guessing game.
Suppose I told you that way back in the year 2000, the two-party vote share for George W. Bush in a particular county was 65%.
Based on this information alone, what would you guess was Donald Trump's two-party vote share there in 2024?

At this point in your political science careers, you might (or might not --- it's ok!) know that a political scientist would often use a linear regression to answer this question.
This entails the following process:

1. Model the relationship between Bush's 2000 vote share and Trump's 2024 vote share as a linear equation: $$\text{Trump2024} \approx \underbrace{\alpha}_{\mathclap{\text{intercept}}} \quad+\quad \underbrace{\beta}_{\mathclap{\text{slope}}} \cdot \text{Bush2000}.$$

2. Estimate the values of the intercept $\alpha$ and the slope $\beta$ that produce the line that most closely fits the data.

3. To come up with the best guess for Trump's 2024 share in a county where Bush got 65% in 2000, plug into the formula: $$\text{Trump2024 estimate} = \hat{\alpha} + \hat{\beta} \cdot 0.65,$$ where $\hat{\alpha}$ and $\hat{\beta}$ are the slope and intercept estimated in step 2.

If we put together the actual data and run a linear regression, we end up with the equation $$\text{Trump2024} \approx 0.17 + 0.86 \cdot \text{Bush2000}.$$
So for our hypothetical county that went 65% for Bush in 2000, we'd guess it went about 73% for Trump in 2024 (via the calculation $0.17 + 0.86 \cdot 0.65$).

```{r} 
#| message: false
#| warning: false
df_pres_county <- df_pres_county |>
  filter(party %in% c("DEMOCRAT", "REPUBLICAN")) |>
  select(year, state = state_po, county = county_name, party, votes = candidatevotes) |>
  # Some counties have separate entries due to apparent FIPS issues?
  group_by(year, state, county, party) |>
  summarize(votes = sum(votes)) |>
  mutate(share = votes / sum(votes)) |>
  ungroup() |>
  filter(party == "REPUBLICAN") |>
  mutate(year = str_c("r_share_", year)) |>
  select(-party, -votes) |>
  pivot_wider(names_from = "year", values_from = "share") |>
  mutate(label = str_c(str_to_title(county), ", ", state)) |>
  filter(!is.na(r_share_2000), !is.na(r_share_2024))
```

```{r}
#| cache: true
#| warning: false
#| label: fig-pres-county-scatter
df_pres_county |>
  ggplot(aes(x = r_share_2000, y = r_share_2024)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_vline(xintercept = 0.65, linewidth = 0.5, color = "red", linetype = "dotted") +
  geom_hline(yintercept = 0.1728 + 0.8605 * 0.65, linewidth = 0.5, color = "red", linetype = "dotted") +
  geom_text_repel(aes(label = label)) +
  labs(
    title = "County-level presidential election returns",
    x = "George W. Bush two-party share, 2000",
    y = "Donald Trump two-party share, 2024"
  )
```

I'll leave it to the stat sequence to explain when and why linear regression is a statistically sound approach to this problem.
We're going to focus on a narrower issue: precisely how we calculate the slope of the regression line, $\hat{\beta}$.
If you open up *Mostly Harmless Econometrics* to page 35, you'll see the "regression anatomy" formula for the slope of a linear regression of dependent variable $Y_i$ on independent variable $X_i$: $$\hat{\beta} = \frac{\text{Cov}[Y_i, X_i]}{V[X_i]} = \frac{\sum_{i=1}^N (X_i - \bar{X}) (Y_i - \bar{Y})}{\sum_{i=1}^N (X_i - \bar{X})^2}.$$
This is indeed the formula that gives us the slope pictured in @fig-pres-county-scatter.

```{r} 
#| echo: true
cov_yx <- cov(df_pres_county$r_share_2024, df_pres_county$r_share_2000)
var_x <- var(df_pres_county$r_share_2000)
cov_yx / var_x
```

But why does this formula "work"?
How did statisticians come up with it?
In what sense, and why, does this particular line fit the data better than any other?
To answer these questions, we will need the tools of [differential calculus]{.concept}.

```{python}
concepts.update({
  "Differential calculus": "Mathematical methods to characterize the rate of change in a function.",
})
```

### Formalizing the problem

For each observation $i = 1, \ldots, N$, we observe an independent variable $X_i$ and a dependent variable $Y_i$, both real numbers.
Given a slope $\beta$, our regression prediction for each observation is $$\hat{Y}_i = \bar{Y} + \beta \cdot (X_i - \bar{X}),$$ where $\bar{Y}$ is the sample mean of $Y_1, \ldots, Y_N$ and $\bar{X}$ is the sample mean of $X_1, \ldots, X_N$.
The residual for the $i$'th observation is the difference between reality and prediction, $$e_i = Y_i - \hat{Y}_i = Y_i - \bar{Y} - \beta \cdot (X_i - \bar{X}).$$

::: {.aside}
Instead of solving for the intercept as well as the slope, I've set up the problem so that the regression line will pass through $(\bar{X}, \bar{Y})$.
Once we get to multivariable calculus, we'll see why it was ok for me to do this.
:::

Substantively, we want our regression line to be as close as possible to the data.
Mathematically, that means we want the residuals to be small in magnitude.
Our specific criterion for the quality of a slope estimate $\beta$ will be the sum of squared residuals when we generate predictions via the slope $\beta$, $$e_1^2 + \cdots + e_N^2 = \sum_{i=1}^N e_i^2 = \sum_{i=1}^N [Y_i - \bar{Y} - \beta (X_i - \bar{X})]^2.$$
Given any slope $\beta$, we can use the formula above to calculate the sum of squared residuals if we drew our regression line according to that slope.
In this sense, the formula sets up a function (see @sec-functions) that maps each possible slope into the associated sum of squared residuals.
Let $f(\beta)$ denote this function: $$f(\beta) = \sum_{i=1}^N [Y_i - \bar{Y} - \beta (X_i - \bar{X})]^2.$$

Finding the best slope is an [optimization problem]{.concept}.
We are looking for the $\beta$ that leads to the lowest sum of squared errors, $f(\beta)$.
Formally, we are looking for a $\hat{\beta}$ that solves $$f(\hat{\beta}) \leq f(\beta) \qquad \text{for all possible values of $\beta$}.$$
Because there are infinite possible values of $\beta$, we cannot solve this problem simply by guessing and checking.
We can --- and will --- use guess-and-check to get close, but we will need calculus to obtain an exact solution.

```{python}
concepts.update({
  "Optimization problem": "The mathematical problem of trying to find the minimal or maximal value of a function.",
})
```

### Approximating the solution

Imagine we didn't know the formula for the regression slope, and we had to figure out what slope best fits the Bush 2000--Trump 2024 regression equation.
We could start by taking a few plausible slopes, calculating the sum of squared residuals for each one, and seeing which one gives us the smallest errors overall.
For example, let's look at each possible slope from 0.0 to 1.0, in intervals of 0.1.

## Limits of functions

### Continuity

## Differentiation

### Derivatives of common functions

### Second derivatives and beyond

## Optimization


```{python, results="asis"}
from helpers import concept_table

print(concept_table(concepts))
```
