---
echo: false
---

# Integrals {#sec-integrals}

```{r setup_integrals}
#| message: false
source("_common.R")
```

## Area under the curve {#sec-area-under-curve}

Let's take a very simple mathematical model of polarized public opinion.
Imagine that everyone has an "ideology," which we'll represent as a real number between $-1$ (most liberal) and $+1$ (most conservative).
There are very many people and very many gradations of ideology, so we will represent the distribution of ideology in the public along a continuum.

To make this a bit more specific, imagine that the [density](https://en.wikipedia.org/wiki/Probability_density_function) of ideology has the following bimodal shape.

```{r}
#| label: fig-ideology-density
#| fig-cap: "A hypothetical bimodal distribution of ideology."
#| cache: true
x <- c(-1, seq(-1, 1, length.out = 201), 1)
y <- x^2 + 1 / 6
y[1] <- y[length(y)] <- 0
tibble(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  scale_x_continuous(breaks = seq(-1, 1, by = 0.5)) +
  scale_y_continuous(breaks = seq(0, 1.25, by = 0.25), limits = c(0, 1.25)) +
  labs(x = "Ideology", y = "Density")
```

Suppose we define a center-left person as someone whose ideology lies between $-1/2$ and $0$.
If the distribution of preferences looks like @fig-ideology-density, then what proportion of people are center left?
To calculate that, we'd want to calculate the area that falls under the curve between $-1/2$ and $0$, and compare that to the total area under the curve.

```{r}
#| label: fig-ideology-center-left
#| fig-cap: "The center-left slice of the ideological distribution."
#| cache: true
df_ideology <- tibble(
  x = seq(-100, 100) / 100, # exact grid, so the filter below hits -1/2 and 0
  y = x^2 + 1 / 6
)
df_ideology |>
  ggplot(aes(x = x, y = y)) +
  geom_ribbon(aes(ymin = 0, ymax = y), fill = mfpa$teal, alpha = 0.08) +
  geom_ribbon(
    aes(ymin = 0, ymax = y),
    data = filter(df_ideology, x >= -1/2, x <= 0),
    fill = mfpa$teal,
    alpha = 0.35
  ) +
  geom_line() +
  annotate("segment", x = c(-1, 1), xend = c(-1, 1), y = 0, yend = 7/6,
           colour = mfpa$teal, linewidth = 1.3) +
  scale_x_continuous(breaks = seq(-1, 1, by = 0.5)) +
  scale_y_continuous(breaks = seq(0, 1.25, by = 0.25), limits = c(0, 1.25)) +
  labs(x = "Ideology", y = "Density")
```

These area calculations are known as [integrals]{.concept entry="Integral" definition="The integral of $f$ between $a$ and $b$, denoted $\int_a^b f(x) \, dx$, is the signed area between the value of $f$ and the x-axis between $x = a$ and $x = b$. (By \"signed\" area, I mean that the value is negative when $f(x) < 0$.)"}.
This type of calculation comes up repeatedly in probability and statistics with continuous random variables, so it is important to at least learn the basics.
We will cover the bare minimum here, just enough for you to get started.
The most important thing to know is a surprising fact---there is a deep connection between derivatives and these area calculations.


## Antiderivatives {#sec-antiderivatives}

It turns out that to calculate the area under a curve given by the function $f(x)$, we will need to find the function whose derivative is equal to $f$.
We call such a function an [antiderivative]{.concept definition="A differentiable function $F(x)$ is an antiderivative of $f(x)$ if $F'(x) = f(x)$ for all $x$."} of $f$.

::: {#def-antiderivative name="Antiderivative"}
Consider functions $f : X \to \mathbb{R}$ and $F : X \to \mathbb{R}$, where $X \subseteq \mathbb{R}$.
$F$ is an [antiderivative]{.concept} of $f$ if $F$ is differentiable and $$F'(x) = f(x)$$ for all $x \in X$.
::: 

As one example, $F(x) = x^2$ is an antiderivative of $f(x) = 2x$.
Using the power rule (see @sec-derivatives-common), we have $$F'(x) = 2 \cdot x^1 = 2x = f(x).$$

::: {#exr-confirm-antiderivative name="Confirming an antiderivative"}
Define the functions $f$, $F$, and $G$, all mapping $\mathbb{R}$ into $\mathbb{R}$, by
$$
\begin{aligned}
f(x) &= 3x^2 + 8x + 7, \\
F(x) &= x^3 + 4x^2 + 7x, \\
G(x) &= x^3 + 4x^2 + 7x + 3.
\end{aligned}
$$
Confirm that $F$ and $G$ are both antiderivatives of $f$.

:::: {.answer}
Using what we know about the derivatives of polynomials (see @exr-deriv-polynomial), we have
$$
\begin{aligned}
F'(x) &= 3x^2 + 4 \cdot 2x^1 + 7 \cdot x^0 \\
&= 3x^2 + 8x + 7 \\
&= f(x); \\
G'(x) &= 3x^2 + 4 \cdot 2x^1 + 7 \cdot x^0 + 0 \\
&= 3x^2 + 8x + 7 \\
&= f(x).
\end{aligned}
$$
::::
::: 

You saw in @exr-confirm-antiderivative that one function may have multiple antiderivatives.
This is not specific to the particular function in the example.
In fact, if a function has one antiderivative, then it has infinitely many antiderivatives: you can form a new one by shifting the existing one up or down by any constant.

::: {#prp-infinite-antiderivatives name="Multiplicity of antiderivatives"}
Let $F$ be an antiderivative of $f$, and let $c \in \mathbb{R}$ be any constant.
The function $G$ defined by $$G(x) = F(x) + c$$ is also an antiderivative of $f$.
:::

::: {.proof}
For all $x$ in the domain, we have
$$
\begin{aligned}
G'(x)
&= \frac{d}{dx} \left[F(x) + c\right] \\
&= F'(x) + \underbrace{\left[\frac{d}{dx} c\right]}_{=0} \\
&= f(x),
\end{aligned}
$$
so $G$ is an antiderivative of $f$.
:::

It is relatively easy to find the antiderivative of a polynomial.
To see why, let's start with the antiderivative of a power function, i.e., a function of the form $f(x) = x^a$.
If $a \neq -1$, then any function of the form $F(x) = \frac{1}{a + 1} x^{a+1} + c$ is an antiderivative.
To see why, use the power rule (see @sec-derivatives-common) and the constant multiple rule (@prp-deriv-constant-multiple):
$$
\begin{aligned}
F'(x)
&= \frac{d}{d x} \left[\frac{1}{a + 1} x^{a + 1} + c\right] \\
&= \frac{1}{a + 1} \cdot \left[\frac{d}{dx} x^{a + 1}\right] + 0 \\
&= \frac{1}{a + 1} \cdot (a + 1) \cdot x^{(a + 1) - 1} \\
&= x^a.
\end{aligned}
$$

Remember that a polynomial is a finite sum of power functions, each multiplied by a coefficient.
Formally, a polynomial is a function of the form $$f(x) = c_n x^n + c_{n-1} x^{n-1} + \cdots + c_1 x + c_0 = \sum_{k=0}^n c_k x^k,$$ where each $c_k$ is a constant real number.
Using the antiderivative of a power function, we have that any function of the form $$F(x) = \frac{c_n}{n + 1} x^{n+1} + \frac{c_{n-1}}{n} x^n + \cdots + \frac{c_1}{2} x^2 + c_0 x + c = \sum_{k=0}^n \frac{c_k}{k + 1} x^{k + 1} + c$$ is an antiderivative of $f$.

::: {#exr-antiderivative-polynomial name="Antiderivatives of polynomials"}
Find an antiderivative of each polynomial listed here.

1. $f(x) = (x + 2) (x - 2)$.
2. $f(x) = 40 x^{39} - 10 x^{17} + 27$.
3. General constant function: $f(x) = \alpha$, where $\alpha$ is a constant.
4. General linear function: $f(x) = \alpha + \beta x$, where $\alpha$ and $\beta$ are constants.
5. General quadratic function: $f(x) = \alpha + \beta x + \gamma x^2$, where $\alpha$, $\beta$, and $\gamma$ are constants.

:::: {.answer}
1. We have $f(x) = x^2 - 2 x + 2 x - 4 = x^2 - 4$, so any function of the form $$F(x) = \frac{x^3}{3} - 4 x + c$$ is an antiderivative of $f$.

2. Any function of the form $$F(x) = x^{40} - \frac{10 x^{18}}{18} + 27 x + c$$ is an antiderivative of $f$.

3. Any function of the form $F(x) = \alpha x + c$ is an antiderivative of $f$.

4. Any function of the form $$F(x) = \alpha x + \frac{\beta x^2}{2} + c$$ is an antiderivative of $f$.

5. Any function of the form $$F(x) = \alpha x + \frac{\beta x^2}{2} + \frac{\gamma x^3}{3} + c$$ is an antiderivative of $f$.
:::: 
:::

::: {#exr-antiderivative-x-negative-one name="Antiderivative of 1/x"}
Why did I need to impose the assumption that $a \neq -1$ when claiming that $F(x) = \frac{1}{a + 1} x^{a + 1} + c$ is an antiderivative of $f(x) = x^a$?
What in fact is the form of the antiderivative of $f(x) = x^{-1}$ for $x \in (0, \infty)$?

:::: {.answer}
We need $a \neq -1$ because otherwise $a + 1 = 0$ and thus $\frac{1}{a + 1}$ is undefined, making the ostensible expression for the antiderivative undefined as well.

Remember that the natural logarithm, $\log(x)$, is defined on the domain $(0, \infty)$ and has a derivative of $\frac{1}{x}$.
Because $\frac{1}{x} = x^{-1}$, we have that any function of the form $F(x) = \log(x) + c$ is an antiderivative of $f(x) = x^{-1}$ on $(0, \infty)$.
::::
:::

## Calculating integrals {#sec-integral-calculation}

### The second fundamental theorem {#sec-second-fundamental-theorem}

Usually I like to lead you through the logic behind a mathematical result *before* telling you the result.
But this result is sufficiently important for day-to-day mathematical life, and the logic behind it is sufficiently intricate, that I'm going to just start with the headline finding.

The [second fundamental theorem of calculus]{.concept definition="If $F$ is an antiderivative of a continuous function $f$, then $\int_a^b f(x) \, dx = F(b) - F(a)$."} states that we can calculate an integral by taking the difference between antiderivatives.

::: {#thm-second-fundamental name="Second fundamental theorem of calculus"}
Let $f : X \to \mathbb{R}$ be continuous, where $X \subseteq \mathbb{R}$, and let $F : X \to \mathbb{R}$ be an antiderivative of $f$.
For any $[a, b] \subseteq X$, we have $$\int_a^b f(x) \, dx = F(b) - F(a).$$
:::

A lot goes into understanding why this theorem "works," so I am going to put that off for @sec-explaining-second-fundamental below.
Before that, let's go through a couple of examples where we can calculate the area separately with normal geometric techniques, and confirm that @thm-second-fundamental gives us the same answer.

Let's start by calculating the area under a constant function, $f(x) = \alpha$, between $a$ and $b$.
As you can see in the illustration below, this amounts to computing the area of a rectangle.

```{r}
#| label: fig-integral-rectangle
#| fig-cap: "Area under a constant function."
#| cache: true
a <- 1
b <- 3
height <- 2 # alpha
df_rect <- tibble(
  x = seq(0, 400) / 100, # exact grid, so the filter below hits a and b
  y = height
)
df_rect |>
  ggplot(aes(x = x, y = y)) +
  geom_ribbon(
    aes(ymin = 0, ymax = y),
    data = filter(df_rect, x >= a, x <= b),
    fill = mfpa$teal,
    alpha = 0.35
  ) +
  geom_line() +
  label_curve(3.5, height + 0.05, 'italic(f)*"(x) = "*alpha', vjust = 0) +
  scale_x_continuous(
    breaks = c(a, b),
    labels = c("a", "b")
  ) +
  scale_y_continuous(breaks = height, labels = expression(alpha)) +
  coord_cartesian(xlim = c(0, 4), ylim = c(0, 3), expand = FALSE) +
  labs(x = "x", y = "f(x)")
```

The width of this rectangle is $b - a$ and its height is $\alpha$, so the area under the curve is $\alpha (b - a)$.
Let's see if the second fundamental theorem gives us the same answer.
Our function is $f(x) = \alpha$, so the function $F(x) = \alpha x$ is an antiderivative.
The second fundamental theorem then gives us
$$
\begin{aligned}
\int_a^b f(x) \, dx
= F(b) - F(a)
= \alpha b - \alpha a
= \alpha (b - a),
\end{aligned}
$$
the same as our geometric area calculation.
So far, so good!

Before moving on to another example, I want to take a second to illustrate why it doesn't matter *which* antiderivative of $f(x)$ we use in the calculation.
Above I used $F(x) = \alpha x$, but in fact any function of the form $F(x) = \alpha x + c$ is an antiderivative of the constant function $f(x) = \alpha$.
But when we plug in another such function, the $c$'s cancel out and we get the same answer:
$$
F(b) - F(a) = [\alpha b + c] - [\alpha a + c] = \alpha (b - a).
$$
This remains true even when calculating integrals of other functions.
So when you have to do such a calculation, just pick whichever antiderivative is easiest (almost always by setting $c = 0$).

Now let's think about calculating the area under the line $f(x) = \beta (x - a)$, from where it crosses the x-axis at $x = a$ up to some other point $x = b$.
As you can see in the illustration below, this amounts to computing the area of a triangle.

```{r}
#| label: fig-integral-triangle
#| fig-cap: "Area under a linear function, starting from where it crosses the x-axis."
#| cache: true
a <- 1
b <- 3
slope <- 1 # beta
df_tri <- tibble(
  x = seq(0, 400) / 100, # exact grid, so the filter below hits a and b
  y = slope * (x - a)
)
df_tri |>
  ggplot(aes(x = x, y = y)) +
  geom_ribbon(
    aes(ymin = 0, ymax = y),
    data = filter(df_tri, x >= a, x <= b),
    fill = mfpa$teal,
    alpha = 0.35
  ) +
  geom_line() +
  label_curve(
    3.25, 2.5, 'italic(f)*"(x) = "*beta*"(x − a)"',
    hjust = 1
  ) +
  scale_x_continuous(
    breaks = c(a, b),
    labels = c("a", "b")
  ) +
  scale_y_continuous(
    breaks = slope * (b - a),
    labels = expression(beta * "(b − a)")
  ) +
  coord_cartesian(xlim = c(0, 4), ylim = c(0, 3.2), expand = FALSE) +
  labs(x = "x", y = "f(x)")
```

The triangle has a width of $b - a$ and a height of $\beta (b - a)$, so its area is $$\frac{1}{2} \cdot (b - a) \cdot [\beta (b - a)] = \frac{\beta}{2} (b - a)^2.$$
Let's again see if @thm-second-fundamental gives us the same answer.
Our function is $f(x) = \beta (x - a) = \beta x - \beta a$, so $F(x) = \frac{\beta}{2} x^2 - \beta a x$ is an antiderivative.
The second fundamental theorem then gives us
$$
\begin{aligned}
\int_a^b f(x) \, dx
&= F(b) - F(a) \\
&= \left[\frac{\beta}{2} b^2 - \beta a b\right] - \left[\frac{\beta}{2} a^2 - \beta a^2\right] \\
&= \frac{\beta}{2} b^2 - \beta a b - \frac{\beta}{2} a^2 + \beta a^2 \\
&= \frac{\beta}{2} b^2 - \beta a b + \frac{\beta}{2} a^2 \\
&= \frac{\beta}{2} \left[b^2 - 2 a b + a^2\right] \\
&= \frac{\beta}{2} (b - a)^2.
\end{aligned}
$$
The algebra to get there was a bit uglier than with the rectangle, but ultimately once again the calculus gives us the same answer as the geometry.

::: {#exr-definite-integral name="Back to the motivating example"}
Let's go back to the example illustrated in @fig-ideology-center-left, where we want to calculate the proportion of people who are "center left."
The function plotted there is the polynomial $$f(x) = x^2 + \frac{1}{6}.$$

1. Identify an antiderivative of $f$.
2. Confirm that the total area under the curve is 1.
3. Find the proportion of people who are center-left by calculating $\int_{-1/2}^0 f(x) \, dx$.

:::: {.answer}
1. One antiderivative is $$F(x) = \frac{x^3}{3} + \frac{x}{6}.$$

2. To calculate the total area under the curve, we take the integral
   $$
   \begin{aligned}
   \int_{-1}^1 f(x) \, dx
   &= F(1) - F(-1) \\
   &= \left[\frac{1}{3} + \frac{1}{6}\right] - \left[\frac{-1}{3} + \frac{-1}{6}\right] \\
   &= 2 \left[\frac{1}{3} + \frac{1}{6}\right] \\
   &= 2 \cdot \frac{3}{6} \\
   &= 1.
   \end{aligned}
   $$

3. The proportion of people who are center-left is
   $$
   \begin{aligned}
   \int_{-1/2}^0 f(x) \, dx
   &= F(0) - F(-1/2) \\
   &= \left[\frac{0}{3} + \frac{0}{6}\right] - \left[\frac{-1/8}{3} + \frac{-1/2}{6}\right] \\
   &= \frac{1}{24} + \frac{1}{12} \\
   &= \frac{3}{24} \\
   &= \frac{1}{8}.
   \end{aligned}
   $$
::::
:::


### Explaining the second fundamental theorem {#sec-explaining-second-fundamental}

TBA.

<!--
::: {.remark title="How deeply do you *need* to understand the second fundamental theorem?"}
I want to admit something here.
I have "known" the second fundamental theorem of calculus, and thus been able to calculate integrals of certain continuous functions, since taking AP Calculus in high school.
Yet I feel like only recently---like, essentially, in the course of having to teach it to you all---have I moved toward *understanding* this theorem.

Most of the things you learn in the methods sequence are things you truly ought to understand at a deep level.
But there are a few results where, for day-to-day life as a working political scientist, it's ultimately okay to understand *that* the result works without fully wrapping your head around *why*.
Over in stats, the [central limit theorem](https://en.wikipedia.org/wiki/Central_limit_theorem) is one of these results.
Here in math, I think the second fundamental theorem of calculus is one of them.

I would love for you to come away with an intuition about rectangular approximations and Riemann sums and their deep connection to the antiderivative.
But at the end of the day, if you just know the formula to use to calculate the integral of a polynomial or other simple continuous function, that's 99% of the battle as a political scientist.
:::
-->

## Properties of the integral {#sec-integral-properties}

Constant multiple rule: $$\int_a^b c f(x) \, dx = c \int_a^b f(x) \, dx.$$

Sum rule: $$\int_a^b [f(x) + g(x)] \, dx = \int_a^b f(x) \, dx + \int_a^b g(x) \, dx.$$

Addition across subdomains: if $a \leq c \leq b$, $$\int_a^b f(x) \, dx = \int_a^c f(x) \, dx + \int_c^b f(x) \, dx.$$

Full statements and explanations TBA.


::: {#concept-review}
:::
